---
title: 'Reproducible Research: Peer Assessment 1'
author: "Ignacio Trujillo"
date: "Sunday, February 15, 2015"
output:
  html_document:
    toc: yes
---

### Basic settings
```{r Basic Settings}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
```

### Loading and preprocessing the data

1. Load the data (i.e. `read.csv()`)

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r Load Data}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destFile <- "activity.zip"
unzippedFile <- "activity.csv"
if(!file.exists(destFile)){download.file(fileUrl,destfile = destFile)}
if(!file.exists(unzippedFile)){unzip(destFile) }
data <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
data$month <- as.numeric(format(data$date, "%m"))
noNA <- na.omit(data)
rownames(noNA) <- 1:nrow(noNA)
```


### What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Make a histogram of the total number of steps taken each day
```{r Histogram of Number of Steps}
library("plyr")
stepsByDate <- ddply(data,~date,summarise,steps=sum(steps))
hist(stepsByDate$steps, 
     col= "steelblue", 
     main = "Histogram of Dialy Steps",
     xlab = "Number of Steps")
```

2. Calculate and report the mean and median total number of steps taken per day

Mean total number of steps taken per day:
```{r Mean of total number of steps}
mean(stepsByDate$steps,na.rm=TRUE)
```
Median total number of steps taken per day:
```{r Median of total number of steps}
median(stepsByDate$steps,na.rm=TRUE)
```

### What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r Daily Average}
stepsByInterval <- ddply(data,~interval,summarise,Steps=mean(steps,na.rm=TRUE))
plot(stepsByInterval$Steps,
     type = "l",
     main = "Average Steps by 5-minute Interval",
     col="steelBlue")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r Maximum Interval}
stepsByInterval[stepsByInterval$Steps == max(stepsByInterval$Steps),]
```

### Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs):

```{r Number of Missing Values}
sum(is.na(data))
```

There are `r sum(is.na(data))` missing values.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

- My strategy is to use the mean for that 5-minute interval to fill each NA value in the steps column.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r Filling missing values}
newData <- data 
for (i in 1:nrow(newData)) {
  if (is.na(newData$steps[i])) {
    newData$steps[i] <- stepsByInterval[which(newData$interval[i] == stepsByInterval$interval), ]$Steps
  }
}

head(newData)
sum(is.na(newData))
```

Now there is `r sum(is.na(newData))` registries with NA values.

4. Make a histogram of the total number of steps taken each day and Calculate and report the *mean* and *median* total number of steps taken per day. 

```{r New Histogram of Number of Steps}
newStepsByDate <- ddply(newData,~date,summarise,steps=sum(steps))
hist(newStepsByDate$steps, 
     col= "steelblue", 
     main = "Histogram of Dialy Steps w/o NAs",
     xlab = "Number of Steps")
```

Mean total number of steps taken per day:
```{r New mean total number of steps}
newTotalSteps <- aggregate(newData$steps, 
                           list(Date = newData$date), 
                           FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean
```
Median total number of steps taken per day:
```{r New median total number of steps}
newMedian <- median(newTotalSteps)
newMedian
```


* Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Compare them with the two before imputing missing data:
```{r Comparison}
oldMean <- mean(stepsByDate$steps,na.rm=TRUE)
oldMedian <- median(stepsByDate$steps,na.rm=TRUE)
newMean - oldMean
newMedian - oldMedian
```

When adding the missing data with the mean of the 5-minute interval, the new mean of total steps is the same but the median became grater that the one without the substitution of NAs.

### Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r Factor into weekday and weekend}
head(newData)
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)
#Since my versión of R is in spanish, I also consider the spanish names of the days
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday","Wednesday","Thursday", "Friday",
                                             "lunes","martes","miércoles","jueves","viernes"),
                                 weekend = c("Saturday", "Sunday",
                                             "sábado","domingo"))
levels(newData$weekdays)
table(newData$weekdays)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r Panel graph}
avgSteps <- aggregate(newData$steps, 
                      list(interval = as.numeric(as.character(newData$interval)), 
                           weekdays = newData$weekdays),
                      FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"
library(lattice)
xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```